---
title: "Hands-on Exercise 2: Choropleth Mapping with R"
execute: 
  warning: false
date: "`r Sys.Date()`"
---

# 2.1 Getting Started

Packages needed:

-   tmap
-   readr for importing delimited text file
-   tidyr for tidying data
-   dplyr for wrangling data
-   sf for handling geospatial data

## 2.1.1 Installing and loading R packages

Installing and loading the packages:

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

-   readr, tidyr, dplyr are part of tidyverse

# 2.2 Importing Data

## 2.2.1 Importing Geospatial Data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

Content of mpsz:

```{r}
mpsz
```

## 2.2.2 Importing Attribute Data

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## 2.2.3 Data Preparation

We need to prepare a data table with year 2020 values before we can prepare a thematic map.

Data Table to include the following variables: PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0 to 4 until age group 20 to 24,
-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,
-   AGED: age group 65 and above,
-   TOTAL: all age group, and
-   DEPENDENCY: the ratio between young and aged against economy active group

### 2.2.3.1 Data wrangling

Data wrangling and transformation functions used:

-   *pivot_wider()* of **tidyr** package
-   *mutate(), filter(), group_by(), select()* of **dplyr** package
