---
title: "In-class Exercise 2: Geospatial Data Wrangling"
execute: 
  message: true
  warning: false
date: "`r Sys.Date()`"
editor: visual
format: html
---

# 2.1 Getting Started

## 2.1.1 Installing and Loading R packages

Install and load sf, tidyverse, funModeling:

```{r}
#| code-fold: true
#| code-summary: Show the code!
pacman::p_load(tidyverse, sf, funModeling)
```

# 2.2 Importing the Data

## 2.2.1 Importing Geospatial Data

Nigeria data from Humanitarian Data Exchange (NGA Data Set):

```{r}
nigeria <- st_read(dsn = "data/geospatial", layer = "nga_polbndl_adm2_1m_salb")
```

geoBoundaries Data Set:

```{r}
geoNGA <- st_read(dsn = "data/geospatial/geoBoundaries", layer = "geoBoundaries-NGA-ADM2") %>% st_transform(crs = 26392)
```

The message shows:

-   polygon feature shapefile with 774 features and 5 fields
-   %\>% is a function called piping from tidyverse to emphasise a sequence of actions rather than the object that the actions are being performed on
-   without the transformation, the geometric data will have very small numbers (in decimal degree)
-   EPSG: [26391](https://epsg.io/26391), 26392 and 26303 are projected coordinate systems for different parts of Nigeria
-   since it is in WGS84, we should do the transformation!

## 2.2.2 Importing Aspatial Data

```{r}
wpdx <- read_csv("data/aspatial/WPdx.csv")
```
