---
title: "In-class Exercise 4: 1st Order Spatial Point Patterns Analysis Methods"
execute: 
  message: true
  warning: false
  echo: true
  eval: true
date: "30 January 2023"
date-modified: "`r Sys.Date()`"
# number-sections: true
editor: visual
format: html
---

# 1.1 Overview

Spatial Point Pattern Analysis is the evaluation of the pattern or distribution, of a set of points on a surface. The point can be location of:

-   events such as crime, traffic accident and disease onset, or
-   business services (coffee and fastfood outlets) or facilities such as childcare and eldercare.

Using appropriate functions of [spatstat](https://cran.r-project.org/web/packages/spatstat/), this hands-on exercise aims to discover the spatial point processes of childecare centres in Singapore.

The specific questions we would like to answer are as follows:

-   are the childcare centres in Singapore randomly distributed throughout the country?
-   if the answer is not, then the next logical question is where are the locations with higher concentration of childcare centres?

# 1.2 The data

To provide answers to the questions above, three data sets will be used. They are:

-   `CHILDCARE`, a point feature data providing both location and attribute information of childcare centres. It was downloaded from Data.gov.sg and is in geojson format.
-   `MP14_SUBZONE_WEB_PL`, a polygon feature data providing information of URA 2014 Master Plan Planning Subzone boundary data. It is in ESRI shapefile format. This data set was also downloaded from Data.gov.sg.
-   `CostalOutline`, a polygon feature data showing the national boundary of Singapore. It is provided by SLA and is in ESRI shapefile format.

# 1.3 Installing and Loading the R packages

In this hands-on exercise, five R packages will be used, they are:

-   [**sf**](https://r-spatial.github.io/sf/), a relatively new R package specially designed to import, manage and process vector-based geospatial data in R.
-   [**spatstat**](https://spatstat.org/), which has a wide range of useful functions for point pattern analysis. In this hands-on exercise, it will be used to perform 1st- and 2nd-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.
-   [**raster**](https://cran.r-project.org/web/packages/raster/) which reads, writes, manipulates, analyses and model of gridded spatial data (i.e. raster). In this hands-on exercise, it will be used to convert image output generate by spatstat into raster format.
-   [**maptools**](https://cran.r-project.org/web/packages/maptools/) which provides a set of tools for manipulating geographic data. In this hands-on exercise, we mainly use it to convert *Spatial* objects into *ppp* format of **spatstat**.
-   [**tmap**](https://cran.r-project.org/web/packages/tmap/) which provides functions for plotting cartographic quality static point patterns maps or interactive maps by using [leaflet](https://leafletjs.com/) API.

Use the code chunk below to install and launch the five R packages.

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap)
```

# 1.4 Spatial Data Wrangling

## 1.4.1 Importing the spatial data

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

### 1.4.2 Mapping the geospatial data sets

```{r}
sf_use_s2(FALSE)
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

```{r}
tmap_mode('plot')
```

# 1.5 Geospatial Data Wrangling

Although simple feature data frame is gaining popularity again sp's Spatial\* classes, there are, however, many geospatial analysis packages require the input geospatial data in sp's Spatial\* classes. In this section, you will learn how to convert simple feature data frame to sp's Spatial\* class.

-   3 steps to converting the simple feature data frame
-   need to do this if u r importing the data using sf function & want to use spatstat
-   1\) convert to sp's spatial class 2) convert into generic sp format 3) convert into spatstat's ppp

## 1.5.1 Converting sf data frames to sp's Spatial\* class

The code chunk below uses [*as_Spatial()*](https://r4gdsa.netlify.app/chap04.html) of **sf** package to convert the three geospatial data from simple feature data frame to sp's Spatial\* class.

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

```{r}
childcare
```

```{r}
mpsz
```

```{r}
sg
```

## 1.5.2 Converting the Spatial\* class into generic sp format

**spatstat** requires the analytical data in ***ppp*** object form. There is no direct way to convert a Spatial\* classes into ***ppp*** object. We need to convert the ***Spatial*** **classes**\* into ***Spatial*** object first.

The codes chunk below converts the Spatial\* classes into generic sp objects.

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

Next, you should display the sp objects properties as shown below.

```{r}
childcare_sp
```

```{r}
sg_sp
```

## 1.5.3 Converting the generic sp format into spatstat's ppp format

Now, we will use *as.ppp()* function of **spatstat** to convert the spatial data into **spatstat**'s **ppp** object format.

```{r}
childcare_ppp <- as(childcare_sp, "ppp")
childcare_ppp
```

Now, let us plot **childcare_ppp** and examine the different.

```{r}
plot(childcare_ppp)
```

You can take a quick look at the summary statistics of the newly created ppp object by using the code chunk below.

```{r}
summary(childcare_ppp)
```

Notice the warning message about duplicates. In spatial point patterns analysis an issue of significant is the presence of duplicates. The statistical methodology used for spatial point patterns processes is based largely on the assumption that process are simple, that is, that the points cannot be coincident.
